---
title: "GSE183947 Bulk RNA-seq (FPKM) Tumor vs Adjacent Normal — Reproducible Analysis Report"
author: "Sophie Tran"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-tools: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

## Overview

This report demonstrates an end-to-end **bulk RNA-seq** analysis workflow using GEO dataset **GSE183947**. The dataset provides a supplementary **FPKM expression matrix** (not raw integer counts). Therefore, differential expression is performed with **limma-trend** on `log2(FPKM + 1)` values (FPKM-appropriate), rather than DESeq2 (which requires raw counts).

### Goals

-   Download and load the GEO supplementary FPKM matrix
-   Construct and verify sample annotations (Tumor vs Normal)
-   Perform QC (distribution, PCA, correlation)
-   Filter low-expression genes
-   Run differential expression (limma-trend)
-   Generate standard outputs (volcano plot, heatmap, marker check)
-   Save figures and tables for sharing/review

------------------------------------------------------------------------

## Setup

```{r}
rm(list = ls())
set.seed(123)
```

## Create output folders

```{r}
dir.create("results", showWarnings = FALSE)
dir.create("results/figures", showWarnings = FALSE)
dir.create("results/tables", showWarnings = FALSE)
```

## Packages

```{r}
pkgs <- c(
  "GEOquery", "dplyr", "tibble", "stringr", "tidyr",
  "ggplot2", "limma", "pheatmap", "matrixStats"
)

to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install) > 0) install.packages(to_install, dependencies = TRUE)

library(GEOquery)
library(dplyr)
library(tibble)
library(stringr)
library(tidyr)
library(ggplot2)
library(limma)
library(pheatmap)
library(matrixStats)
```

## Data acquisition (GEO supplementary file)

```{r}
gse_id <- "GSE183947"

# Download supplementary files into folder ./GSE183947/
getGEOSuppFiles(gse_id, makeDirectory = TRUE)

fpkm_path <- file.path(gse_id, paste0(gse_id, "_fpkm.csv.gz"))
if (!file.exists(fpkm_path)) {
  stop("FPKM file not found. Check list.files('GSE183947') for exact filename.")
}

counts_raw <- read.csv(
  fpkm_path,
  row.names = 1,
  check.names = FALSE
)

counts_raw <- as.matrix(counts_raw)
storage.mode(counts_raw) <- "numeric"

dim(counts_raw)
```

### Notes on input data

-   This file contains **FPKM** values (continuous, normalized) rather than raw counts.

<!-- -->

-   Because of this, we use **limma-trend** (linear modeling + empirical Bayes) on `log2(FPKM+1)`

## Sample annotation (Tumor vs Normal)

The column naming convention in the FPKM matrix indicates sample groups:

-   CA.\* = Tumor (Cancer)

-   CAP.\* = Adjecent/normal tissue (Cancer Adjecent Paratumor) -\> treated as Normal here

    ```{r}
    sample_info <- data.frame(
      sample_id = colnames(counts_raw),
      condition = ifelse(grepl("^CA\\.", colnames(counts_raw)), "Tumor",
                         ifelse(grepl("^CAP\\.", colnames(counts_raw)), "Normal", NA_character_))
    )

    if (any(is.na(sample_info$condition))) {
      bad <- sample_info %>% filter(is.na(condition))
      stop("Some samples did not match CA./CAP. pattern:\n", paste(bad$sample_id, collapse = ", "))
    }

    sample_info$condition <- factor(sample_info$condition, levels = c("Normal", "Tumor"))
    rownames(sample_info) <- sample_info$sample_id

    table(sample_info$condition)
    ```

```{r}
write.csv(sample_info, "results/tables/sample_info.csv", row.names = TRUE)
```

## QC (FPKM-aware)

### 1) Sample-level sum of FPKM (not true library size)

Because values are FPKM, the sum across genes is not a sequencing library size. However, it can still flag samples with unusual overall expression magnitude.

```{r}
sample_totals <- colSums(counts_raw, na.rm = TRUE)
qc_df <- data.frame(
  sample_id = names(sample_totals),
  total_fpkm_sum = as.numeric(sample_totals),
  condition = sample_info[names(sample_totals), "condition"]
) %>% arrange(desc(total_fpkm_sum))

p_totals <- ggplot(qc_df, aes(x = reorder(sample_id, total_fpkm_sum),
                              y = total_fpkm_sum / 1e6,
                              fill = condition)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "QC: Sum(FPKM) per sample (not true library size)",
    x = NULL,
    y = "Sum(FPKM) (millions)",
    fill = "Condition"
  ) +
  theme_minimal(base_size = 12)

p_totals
```

```{r}
ggsave("results/figures/qc_sum_fpkm_per_sample.png", p_totals, width = 10, height = 7, dpi = 150)
```

### 2) Distribution of log2(FPKM+1)

```{r}
expr_log <- log2(counts_raw + 1)

boxplot(expr_log,
        outline = FALSE,
        las = 2,
        main = "QC: log2(FPKM+1) distribution per sample",
        ylab = "log2(FPKM + 1)")
```

## Filtering low-expression genes

Criterion used:

-   Keep genes with FPKM \> 1 at least 5 samples

```{r}
keep <- rowSums(counts_raw > 1) >= 5
cat("Genes before filtering:", nrow(counts_raw), "\n")
cat("Genes after filtering :", sum(keep), "\n")

counts_f <- counts_raw[keep, , drop = FALSE]
expr_f <- log2(counts_f + 1)
```

## PCA (sample QC/ group separation)

```{r}
pca <- prcomp(t(expr_f), scale. = TRUE)

pca_df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  sample_id = rownames(pca$x),
  condition = sample_info[rownames(pca$x), "condition"]
)

p_pca <- ggplot(pca_df, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 3, alpha = 0.9) +
  labs(title = "PCA of log2(FPKM+1) (filtered genes)") +
  theme_minimal(base_size = 12)

p_pca
```

```{r}
ggsave("results/figures/qc_pca.png", p_pca, width = 8, height = 6, dpi = 150)
```

## Sample correlation heatmap

```{r}
cor_mat <- cor(expr_f, method = "pearson")

ann_col <- data.frame(condition = sample_info$condition)
rownames(ann_col) <- rownames(sample_info)

pheatmap(cor_mat,
         annotation_col = ann_col,
         show_rownames = FALSE,
         show_colnames = FALSE,
         main = "Sample correlation heatmap (log2 FPKM+1)")
```

```{r}
png("results/figures/qc_sample_correlation_heatmap.png", width = 1100, height = 900, res = 150)
pheatmap(cor_mat,
         annotation_col = ann_col,
         show_rownames = FALSE,
         show_colnames = FALSE,
         main = "Sample correlation heatmap (log2 FPKM+1)")
dev.off()
```

## Differential expression (limma-trend)

-   Reference level: Normal

-   Coefficient: conditionTumor (Tumor vs Normal)

    ```{r}
    design <- model.matrix(~ condition, data = sample_info)

    fit <- lmFit(expr_f, design)
    fit <- eBayes(fit, trend = TRUE)

    res <- topTable(fit, coef = "conditionTumor", number = Inf, sort.by = "P")
    res$gene <- rownames(res)

    head(res[, c("gene", "logFC", "AveExpr", "P.Value", "adj.P.Val")], 10)
    ```

```{r}
write.csv(res, "results/tables/limma_DE_Tumor_vs_Normal.csv", row.names = FALSE)
```

## Volcano plot

```{r}
res2 <- res %>%
  mutate(
    neglog10FDR = -log10(adj.P.Val + 1e-300),
    sig = adj.P.Val < 0.05 & abs(logFC) >= 1
  )

p_volcano <- ggplot(res2, aes(x = logFC, y = neglog10FDR)) +
  geom_point(aes(shape = sig), alpha = 0.6) +
  labs(
    title = "Volcano plot (limma-trend on log2(FPKM+1))",
    x = "log2 fold change (Tumor vs Normal)",
    y = "-log10(FDR)"
  ) +
  theme_minimal(base_size = 12)

p_volcano
```

```{r}
ggsave("results/figures/de_volcano.png", p_volcano, width = 8, height = 6, dpi = 150)
```

## Heatmap of top 50 DE genes

```{r}
top50 <- res %>%
  arrange(adj.P.Val) %>%
  slice(1:50) %>%
  pull(gene)

mat_top <- expr_f[top50, , drop = FALSE]

# Z-score by gene for visualization
mat_top_z <- t(scale(t(mat_top)))

pheatmap(mat_top_z,
         annotation_col = ann_col,
         show_colnames = FALSE,
         fontsize_row = 8,
         main = "Top 50 DE genes (z-scored, log2 FPKM+1)")
```

```{r}
png("results/figures/de_heatmap_top50.png", width = 1100, height = 1200, res = 150)
pheatmap(mat_top_z,
         annotation_col = ann_col,
         show_colnames = FALSE,
         fontsize_row = 8,
         main = "Top 50 DE genes (z-scored, log2 FPKM+1)")
dev.off()
```

## Canonical marker check

```{r}
markers <- c("ESR1", "ERBB2", "MKI67", "TP53", "KRT19")

marker_res <- res %>%
  dplyr::filter(gene %in% markers) %>%
  dplyr::select(gene, logFC, AveExpr, P.Value, adj.P.Val)

marker_res
```

```{r}
write.csv(marker_res, "results/tables/marker_check.csv", row.names = FALSE)
```

Interpretation guidance:

-   LogFC \> 0 -\> higher in Tumor

-   LogFC \< 0 -\> higher in Normal

## Notes on limitations

-   This analysis begins from an FPKM matrix supplied by GEO, so upstream steps (alignment/quantification from FASTQ) are not reproduced here.

-   For raw-count workflows, tools such as STAR/Hisat2 + featureCounts/HTSeq + DESeq2 would be appropriate.

## Key findings

### Differential Expression Summary

```{r}
# Number of significant genes
n_sig <- sum(res$adj.P.Val < 0.05)
n_up  <- sum(res$adj.P.Val < 0.05 & res$logFC > 0)
n_dn  <- sum(res$adj.P.Val < 0.05 & res$logFC < 0)

summary_df <- data.frame(
  Metric = c("Total genes tested",
             "Significant genes (FDR < 0.05)",
             "Upregulated in Tumor",
             "Higher in Normal"),
  Value = c(nrow(res), n_sig, n_up, n_dn)
)

summary_df
```

### Top 10 Upregulated Genes in Tumor

```{r}
top_up <- res %>%
  dplyr::filter(adj.P.Val < 0.05, logFC > 0) %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, logFC, adj.P.Val) %>%
  head(10)

top_up
```

### Top 10 Genes Higher in Normal Tissue

```{r}
top_dn <- res %>%
  dplyr::filter(adj.P.Val < 0.05, logFC < 0) %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, logFC, adj.P.Val) %>%
  head(10)

top_dn
```

### Canonical Breast Cancer Marker Summary

```{r}
marker_res
```

## Interpretation

```{r}
cat("
This bulk RNA-seq analysis identified", n_sig,
"genes significantly differentially expressed (FDR < 0.05) between tumor and adjacent normal tissue.

Of these,",
n_up, "genes were upregulated in tumor samples, while",
n_dn, "genes showed higher expression in normal tissue.

Upregulated tumor genes include proliferation-associated and epithelial markers, 
consistent with malignant transformation and increased cellular proliferation.

Reduced expression of hormone receptor markers in tumor samples suggests 
potential subtype heterogeneity or reduced hormone receptor signaling in this cohort.
")
```
